---
title: "Lab 06: Statistical Inference in Multiple Linear Regression - Part 1"
subtitle: "Econometrics I - YTU"
author: 
  name: "Prof. Dr. Hüseyin Taştan"
  affiliation: "Yıldız Technical University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: 2020 Fall
output: 
  html_document:
    number_sections: true
    theme: readable
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---
<style type="text/css"> 
body{
  font-size: 12pt;
}
code.r{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'asis')
``` 


# t-test Examples

## Wage Equation

The model is given by 
$$
\log(wage) = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 tenure + u
$$
Estimate the model using `wage1 ` data. 
```{r, echo=TRUE, results="hold"}
library(wooldridge)
wagereg <- lm(log(wage) ~ educ + exper + tenure, data = wage1)
summary(wagereg) 
``` 
Test the statistical significance of the coefficient on exper against one-sided alternative, i.e., 
$$
H_0: \beta_2 = 0
$$
$$
H_1: \beta_2 > 0
$$

The t-statistic is already reported in the output, $t_{\hat{\beta}_2}=2.391$. We need 
to compare this to the critical value from the t distribution with dof=522 degrees of freedom. Let's find the critical values at 5% and 1% significance levels using R:  
```{r, echo=TRUE}
# CV for alpha=5% and 1% using the t distribution with 522 d.f.
# Using the quantile function for t (qt)
alpha <- c(0.05, 0.01)
qt(1-alpha, 522)

# Critical values for alpha=5% and 1% using the normal approximation:
# Using the quantile function for normal distribution (qnorm)
qnorm(1-alpha)
``` 

Recall that for large degrees of freedom (approximately dof>120), the normal approximation gives practically the same answer. This is no surprise because as the sample size increases (assuming the number of parameters remain the same), t distribution approaches the standard normal distribution. Thus, for large samples we can just use the standard normal critical values. 

Because the t test statistic is larger than the critical value, $t_{\hat{\beta}_2}=2.391>2.33$, at 1% significance level, we reject the null hypothesis in favor of the alternative. Although the coefficient estimate is small, it is statistically different from zero. 

We can find the p-value using the distribution function of t in R (pt): 
```{r, echo=TRUE}
pt(2.391,522, lower.tail = FALSE)
# or 
1-pt(2.391,522)
``` 

Note that this is half the p-value reported in the regression output. 
Because p-value<0.01, we can reject the null hypothesis at 1% significance level.

## Student Performance Model, level-level

```{r, echo=TRUE, results="hold"}
library(wooldridge)
sumres1 <- summary(lm(math10 ~ totcomp + staff + enroll, data = meap93))
regtable1 <- sumres1$coefficients
regtable1
``` 

Manual confirmation of test statistics: 
```{r, echo=TRUE, results="hold"} 
bhat <- regtable1[,1]
se   <- regtable1[,2]

# Reproduce t statistic
( tstat <- bhat / se ) 
``` 
```{r, echo=TRUE, results="hold"}
# Reproduce p value
( pval  <- 2*pt(-abs(tstat),404) )
``` 

## Student Performance Model, level-log
```{r, echo=TRUE, results="hold"}
sumres2 <- summary (lm(math10 ~ log(totcomp) + log(staff) + log(enroll), data = meap93))
regtable2 <- sumres2$coefficients
regtable2
``` 

**Exercise**: Is log(enroll) statistically significant? Test this null hypothesis against the lower tail alternative hypothesis. Find the p-value and state your decision. 

## College GPA 
```{r, echo=TRUE, results="hold"}
( sumres <- summary( lm(colGPA ~ hsGPA+ACT+skipped, data=gpa1) ) )
``` 
```{r, echo=TRUE, results="hold"}
# Manually confirm the formulas: Extract coefficients and SE
regtable <- sumres$coefficients
bhat <- regtable[,1]
se   <- regtable[,2]

# Reproduce t statistic
( tstat <- bhat / se )
# Reproduce p value
( pval  <- 2*pt(-abs(tstat),137) )
``` 

## Campus Crime Model 
```{r, echo=TRUE, results="hold"}
( crimeres <- summary( lm(log(crime) ~ log(enroll), data=campus) ) )
``` 

## House prices 
```{r, echo=TRUE, results="hold"}
( houseres <- summary( lm(log(price) ~ log(nox)+log(dist)+rooms+stratio, data=hprice2) ) )
``` 

# Confidence Interval Example: R&D Expenditures 
$$
\log (\mathrm{rd})=\beta_{0}+\beta_{1} \cdot \log (\mathrm{educ})+\beta_{2} \cdot \mathrm{profmarg}+u
$$
```{r, echo=TRUE, results="hold"}
# OLS regression:
myres <- lm(log(rd) ~ log(sales)+profmarg, data=rdchem)

# Regression output:
summary(myres)
``` 

```{r, echo=TRUE, results="hold"}
# 95% CI:
confint(myres)
```
```{r, echo=TRUE, results="hold"}
# 99% CI:
confint(myres, level=0.99)
``` 



<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>


