---
title: "Lab 13: Misspecification and Data Problems"
subtitle: "Econometrics I - YTU"
author: 
  name: "Prof. Dr. Hüseyin Taştan"
  affiliation: "Yıldız Technical University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: 2020 Fall
output: 
  html_document:
    number_sections: true
    theme: readable
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---
<style type="text/css"> 
body{
  font-size: 12pt;
}
code.r{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hold', fig.show = 'asis')
``` 

# Functional Form Misspecification  
 
We can use Ramsey's RESET test to detect functional form misspecification. 

## RESET Test Example:  Housing Price Model

Automatic computation of RESET test using `lmtest` package: 

```{r, echo=TRUE, eval=TRUE}
library(wooldridge)
# don't forget to install lmtest before you load them
library(lmtest)  
# original linear regression
orig <- lm(price ~ lotsize + sqrft + bdrms, data = hprice1)

# RESET test
library(lmtest)
resettest(orig)
```


Manual computation of RESET test: 

```{r, echo=TRUE, eval=TRUE}
# original linear regression
orig <- lm(price ~ lotsize + sqrft + bdrms, data = hprice1)

# regression for RESET test
RESETreg <- lm(price ~ lotsize + sqrft + bdrms + 
                 I(fitted(orig)^2) + I(fitted(orig)^3), data=hprice1)
# note that we added powers of fitted values
RESETreg

# RESET test. H0: all coeffs including "fitted" are=0 
library(car)
linearHypothesis(RESETreg, matchCoefs(RESETreg,"fitted"))
```


RESET F statistic is 4.67 with p-value=0.01. Thus, the null of no functional form misspecification is rejected at 5% level. 

Use logarithmic functional form: 

```{r, echo=TRUE, eval=TRUE}
logreg <- lm(log(price) ~ log(lotsize) + log(sqrft) + bdrms, data = hprice1)

# RESET test
library(lmtest)
resettest(logreg)

```

Interpret the RESET test result. Use 5% statistical significance level. 

# Non-nested tests 

Consider two house price models, model1 and model2, which are non-nested alternatives. 
The model that includes both model1 and model2 as special cases (i.e., nested) is called the **encompassing** model. 


```{r, echo=TRUE, eval=TRUE}
# two alternative models
model1 <- lm(price ~     lotsize  +     sqrft  + bdrms, data = hprice1)
model2 <- lm(price ~ log(lotsize) + log(sqrft) + bdrms, data = hprice1)

# Test against comprehensive model
library(lmtest)
encomptest(model1, model2, data = hprice1)

```

According to the encompassing test both models are rejected against the encompassing model. (note that the dependent variable must be the same in all models).

# Missing Data 
In many data sets, the existence of missing observations is usual. In R, missing data can be represented by different values.
__NA__ (not available) refers to no information.
__NaN__ (not a number) indicates that the value is not defined.
The function __is.na()__ returns _TRUE_ if the observation is either __NA__ or __NaN__ and _FALSE_ otherwise. 

Operaitons resulting in $\pm\infty$ like $log(0)$ or $1/0$ are not coded as __NaN__ but as __Inf__ or __-Inf__. R can do calculations with these numbers.

```{r, echo=TRUE, results='hold'}
x <- c(-1,0,1,NA,NaN,-Inf,Inf,9999)
logx <- log(x)
invx <- 1/x
ncdf <- pnorm(x)
isna <- is.na(x)
mydata<-data.frame(x, logx, invx, ncdf, isna)
mydata
```

Some data may include unusual observations. Assume that _9999_ is an impossible value which may be replaced by __NA__ with the help of following command.

```{r, echo=TRUE, results='hold'}
mydata[mydata==9999]<-NA
mydata
```
It may be useful to count the missing observations. __colSums(is.na())__ command counts the missings in each variable while __complete.cases()__ generates a logical vector indicating the rows of the data frame which don't have any missing information.
```{r, echo=TRUE, results='hold'}
# Frequencies of the missings.
table(is.na(mydata))
# Number of the missings in each variable.
colSums(is.na(mydata))
# Information of complete cases by row.
complete.cases(mydata)
table(complete.cases(mydata))
```
Although there are advanced methods to deal with missing data a widely used strategy is to remove the observations with missing values. __lm__ command does this by default.
```{r, echo=TRUE, results='hold'}
data(lawsch85, package='wooldridge')

# Mean of a variable with missings:
mean(lawsch85$LSAT)
mean(lawsch85$LSAT, na.rm=TRUE)

# Regression with missings
summary(lm(log(salary) ~ LSAT + cost + age, data = lawsch85))
```
# Outliers 

Example: a model of R&D expenditures (see classnotes)

```{r, echo=TRUE, eval=TRUE}
library(wooldridge)
data("rdchem")
reg <- lm(rdintens ~ sales + profmarg, data = rdchem)
plot(rdchem$sales, rdchem$rdintens)

# Studentized residuals for all observations:
studres <- rstudent(reg)

# Display extreme values:
min(studres)
max(studres)

# Histogram (and overlayed density plot):
hist(studres, freq=FALSE)
lines(density(studres), lwd=2)
```

Logarithmic Regression: 

```{r, echo=TRUE, eval=TRUE}
library(wooldridge)
data("rdchem")
reg <- lm(log(rd) ~ log(sales) + profmarg, data = rdchem)
plot(log(rdchem$sales), log(rdchem$rd))


# Studentized residuals for all observations:
studres <- rstudent(reg)

# Display extreme values:
min(studres)
max(studres)

# Histogram (and overlayed density plot):
hist(studres, freq=FALSE)
lines(density(studres), lwd=2)
```





<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>


